---
title: "Titanic_Survival"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Titanic Survival 

In this notebook , I'll try to analyse who had the greater chance of survival and try to answer those question.
And finaly if we find a pattern , We'll build a classification model to predict who survived|Died.



```{r}
  
library("tidyverse")

train <- read_csv("C:/Users/Alaa/Desktop/train.csv")
test <- read_csv("C:/Users/Alaa/Desktop/test.csv")

data <- bind_rows(train,test)


```



## Let's explore our data and look for interesting variables !


How many levels in the variables Embarked we have ? And what's the proportion ? 

```{r}
data %>%
  count(Embarked,sort = TRUE)
```

* Here are the levels are place of emabarkement :
    + *S:Southampton*
    + *C:Cherbourg*
    + *Q:Queenstown*
 
 We also have *two NA* which we'll try and deal with them later.
 
 
 What about **Pclass**
 
```{r}
data %>%
  count(Pclass,sort = TRUE)
```
 

We have three classes , mostly people are in the third class . 
We can see there is a relationship between class and Fare (price of the ticket)


How is the Variable **Fare** distributed ? 

```{r}
data %>%
  ggplot(mapping = aes(x= PassengerId , y = Fare , color = as.factor(Pclass )))+
  geom_point(size = 2,alpha = 0.7)+
  theme_minimal()+
  labs(color = "PClass" , title = "Relationship between fare and Pclass")
  
```


Clearly the lower the classes the higher the fare 

I see some value equal to 0 , let's check this value in our data

```{r}
data %>%
  filter(Fare == 0 |is.na(Fare)) %>%
  view()

```

Cool ! All passenger with fare equal are mostly crew of the titanic , other working for the AMERICAN LINE ... 
And got free ticket because their trip were canceled with philadephia ship.

```{r}
data %>%
  mutate(statut = ifelse(Fare == 0|is.na(Fare) , "employee" , "passenger")) -> data



train_p <- data[1:891,] 

train_p %>%
  filter(Fare == 0 |is.na(Fare)) %>%
  view()
train_p %>%
  group_by(statut)%>%
  summarise(ratio_dying = 1-  sum(Survived)/n()) ->statut_survival 

statut_survival%>%
  
  ggplot(mapping = aes(x = statut , y = ratio_dying , color = statut))+
  geom_point(size = 3) + 
  theme_minimal()+
  geom_segment(aes(x = statut , y = 0 , xend = statut , yend = ratio_dying),size = 2)+
  scale_y_continuous(label  =scales::percent_format()) + 
  labs(y = "Probaility of dying")
  

```






Let's again do some graphic with some Boxplot


```{r}
data %>%
  ggplot(mapping = aes(x= as.factor(Pclass), y = Fare , fill = as.factor(Pclass )))+
  geom_boxplot()+
  theme_minimal()+
  labs(fill = "PClass",x = "Pclass")+
  scale_y_continuous(label = scales::label_dollar())

```


Is there a relationship between survival and PClass ?


```{r}
train_p %>%
  ggplot(mapping = aes(x = Pclass , fill = as.factor(Survived)))+
  geom_bar(stat = "count")+
  coord_flip()+
  theme_minimal()
  
```

We can see that most Passenger that died were 3rd class passenger ! And the most who survived were 1rst class passenger  !


Back in the boxplot the Pclass = 1 , had a lot of variation when it comes to fare prices .
We'll see if the price affect the survival rate


```{r}
train_p %>%
  filter(Pclass == 1)%>%
  summary()  #The median of the variable Fare is 60.29

train_p %>%
  filter(Pclass ==1) %>%
  mutate(Price_Pclass = case_when(Fare < 60.29 ~ "below_median",
                                  Fare >=60.29 ~ "top_median"))%>%
  ggplot(mapping = aes(x = Price_Pclass , fill  =as.factor(Survived)))+
  geom_bar(stat = "count")+
  theme_minimal()
```

There is a difference ! What if we did more ? let's try and didive it in 4 . 

```{r}
train_p %>%
  filter(Pclass ==1) %>%
  mutate(Price_Pclass = case_when(Fare < 30.92 ~ "quantile_1",
                                  Fare >=30.92 & Fare < 60.29 ~ "quantile_2",
                                  Fare >=60.29 & Fare < 93.5 ~"quantile_3",
                                  Fare >=93.5 ~ "quantile_4"))%>%
  ggplot(mapping = aes(x = Price_Pclass , fill  =as.factor(Survived)))+
  geom_bar(stat = "count")+
  theme_minimal()+
  coord_flip()+
  labs(fill = "Survived" , x = "Quantile" , title = "Quantitle of variable fare for the passenger of 1rst class") 
```
Let's add this price_pclass to our data

```{r}
data %>%
    mutate(Price_Pclass = case_when(Pclass==1 & Fare < 30.92 ~ "Pclass1_4",
                                  Pclass==1 & Fare >=30.92 & Fare < 60.29 ~ "Pclass1_3",
                                  Pclass==1 & Fare >=60.29 & Fare < 93.5 ~"Pclass1_2",
                                  Pclass==1 & Fare >=93.5 ~ "Pclass1_1",
                                  TRUE ~ as.character(Pclass))) -> data
```


Hmm this is sad , but it seems once again money buy survival  , or maybe this had to do with the location of the cabin ?  

Can we do better than the quantile ? 

```{r}
data %>%
  count(Fare,sort = TRUE)%>%
  head(n= 15) %>%
  inner_join(data,by ="Fare")%>%
  group_by(Fare)%>%
  summarise(ratio = sum(Survived,na.rm = TRUE)/n())%>%
  ungroup()%>%
  mutate(Fare = as.factor(Fare),
         Fare = fct_reorder(Fare,ratio,sum))%>%
  ggplot(mapping = aes(x = as.factor(Fare) , y = ratio ,fill =as.factor(Fare)))+
  theme_minimal()+
  geom_bar(stat = "identity")+
  coord_flip()+
  theme(panel.grid = element_blank(),
        legend.position = "none")+
  labs(x ="Fare as a factor variable")#Graphic Fare



data%>%
  filter(Fare == 512.3292 | Fare ==263.0000 & Sex =="female") %>%
  view()
```

There is one exception in the graphic fare the fare == 7.925 has hight rate of survival and I wonder why ???
Let's take a look


```{r}
data %>%
  filter(Fare == 7.925)%>%
  view() #Occupation == General Labourer
        # Nationnality mostly finnish

data %>%
  filter(Fare == 26.55)%>%
  view() #businessman ,lawyer,journalist ....


data %>%
  filter(Fare == 7.75)%>%
  view() #nationality Irsih 
         #occuption mostly farmer

data %>%
  filter(Fare == 13)%>%
  view() # E D F ? 
         #Nationnality english danish polish ...


data %>%
  filter(Fare == 7.2292)%>%
  view() #Lebanese =) 



data %>%
  filter(Fare == 7.8958)%>%
  view() #bulgarian

data %>%
  count(Fare,sort = TRUE)%>%
  filter(n ==4) %>%
  inner_join(data,by ="Fare")%>%
  view()


data %>%
  group_by(Fare,Ticket)%>%
  summarise(count = n())%>%
  filter( count >= 2) %>%
  arrange((desc(count)))%>%
  view()

data %>%
  group_by(Ticket)%>%
  mutate(Linked_person = n(),
         Linked_person = as.factor(Linked_person)) %>%
  ungroup() ->data
  

data %>%
  filter(!is.na(Survived))%>%
  ggplot(mapping = aes(x = Linked_person , y= as.factor(Survived) , fill = as.factor(Survived)))+
  geom_bar(stat = "identity")+
  theme_minimal()+
  facet_wrap(~Sex)+
  coord_flip() 
  

```


```{r}

```





Let's look at the ticket variable and see if we can find something interesting
```{r}
data %>%
  count(Ticket,sort = TRUE)%>%
  head()



```


After checking the code below , it seems female in the Pclass with expensive had very high chance of survival , and male not the same case . The only explanation is that there was no exception no matter where you came from it was women
and children first , but male in the first class were making sure that their family (women and children) were on the
lifeboat.


Let's look at the lass graphic but with sex as factor now also.

```{r}

train_p %>%
  filter(Pclass ==1) %>%
  mutate(Price_Pclass = case_when(Fare < 30.92 ~ "quantile_1",
                                  Fare >=30.92 & Fare < 60.29 ~ "quantile_2",
                                  Fare >=60.29 & Fare < 93.5 ~"quantile_3",
                                  Fare >=93.5 ~ "quantile_4"))%>%
  ggplot(mapping = aes(x = Price_Pclass , fill  =as.factor(Survived)))+
  geom_bar(stat = "count")+
  theme_minimal()+
  coord_flip()+
  labs(fill = "Survived" , x = "Quantile" , title = "Quantitle of variable fare for the passenger of 1rst class") +
  facet_wrap(~Sex)
```

I want to check the same graph but also with Age as a variable

```{r}
train_p %>%
  filter(Pclass ==1    & !is.na(Age)) %>%
  mutate(Price_Pclass = case_when(Fare < 30.92 ~ "quantile_1",
                                  Fare >=30.92 & Fare < 60.29 ~ "quantile_2",
                                  Fare >=60.29 & Fare < 93.5 ~"quantile_3",
                                  Fare >=93.5 ~ "quantile_4"),
         age_category =  case_when(Age < 16 ~"child",
                                   Age >= 16 ~ "adult")) -> pclass1_children
  pclass1_children%>%
  ggplot(mapping = aes(x = Price_Pclass , fill  =as.factor(Survived)))+
  geom_bar(stat = "count")+
  theme_minimal()+
  coord_flip()+
  labs(fill = "Survived" , x = "Quantile" , title = "Quantitle of variable fare for the passenger of 1rst class") +
  facet_wrap(~Sex+age_category)
```

We have a lot of **NA** in the age variable so we can't really make a conclusion here .


```{r}
pclass1_children %>%
  filter(age_category == "child" & Sex == "female")%>%
  view()
```


The girl who didn't survive was a baby =( this means her mother didn't also survive


Let's check the distribution of age 





```{r}
data %>%
  filter(!is.na(Age))%>%
  ggplot(mapping = aes(x  = Age ))+
  geom_histogram(fill  = "#f1c40f")+
  theme_minimal()+
  labs(title = "Distribution of the Age variable")


data %>%
  filter(!is.na(Age))%>%
  ggplot(mapping = aes(x = Age))+
  geom_density()+
  theme_minimal()



```
I want to transform my age variable to a category variable , but before doing that I need to fix the NA . 
I'm going to use the well known mice package for that


```{r}
library("mice")

md.pattern(data)


mice_pred <- mice(data = data[,!names(data) %in% c("PassengerId","Name","Survived","Cabin" ,"Fare")], method = "rf")

mice_output <- complete(mice_pred)

```


Let's check if the density is still the same ! 



```{r}
data %>%
  ggplot(mapping = aes(x = Age))+
  geom_density(color = "blue", size = 1.5 )+
  geom_density(aes(x = mice_output$Age ),color = "red",size = 1.5)+
  theme_minimal()
```
That's great !!! The distribution hasn't changed 

```{r}
data$Age <- mice_output$Age
```

Fixing the Embarked NA value
```{r}
data %>%
  filter(is.na(Embarked))%>%
  view()
```
I googled and they both come from Southampton , suppose we didn't have google or another source we could have guess
by the cabin variable , they were in cabin B28

```{r}
data %>%
  filter(grepl("B2",Cabin))%>%
  ggplot(mapping = aes(x = Embarked, fill = Embarked))+
  geom_bar(stat = "count")+
  theme_minimal()



data$Embarked[is.na(data$Embarked)] <- "S"
```

```{r}

data %>%
  mutate(Embarked_Pclass = paste0(Embarked,"_",Price_Pclass)) ->data
```




##Prediction 


```{r}
library("randomForest")

data %>%
  mutate_if(is.character,factor)%>%
  mutate(Survived = as.factor(Survived))->data

train_p <- data[1:891,]
test_p <- data[892:1309,]
str(train_p)


model_rf <-randomForest(Survived ~ Age + Sex  + statut + Linked_person + Embarked_Pclass , data = train_p)

prediction <- predict(model_rf,test_p)

solutions2 <- data.frame(PassengerId = test_p$PassengerId,Survived = prediction)

write.csv(solutions2 , file = "solutions_3.csv", row.names = FALSE)

```


Let's analyse the age variable


```{r}
summary(data$Age)

data %>%
  filter(!is.na(Survived))%>%
  group_by(Age)%>%
  mutate(count = n())%>%
  filter(count >= 5) %>%
  summarise(ratio_age = sum(Survived)/n())%>%
  ggplot(mapping = aes(x= Age , y= ratio_age))+
  geom_line()+
  theme_minimal()+
  scale_x_continuous(breaks = seq(0,50,2))+
  expand_limits(y =0 ,x = 0) #Hard to see any difference

#0 - 4 age_1
#5-18 - age_2
#19 - 35 
#35 -50 
#50 - 60
#60+


data %>%
  mutate(age_category = case_when(Age <= 4 ~ "age_1",
                                  Age >4 & Age <=18 ~ "age_2",
                                  Age>18 & Age <40 ~  "age_3",
                                  Age >= 40 ~"age_4")) ->data
  data %>%
  filter(!is.na(Survived))%>%
    mutate(age_category = fct_reorder(age_category , as.numeric(Survived),mean))%>%
  ggplot(mapping = aes(x = age_category , fill =as.factor(Survived) ) )+
  geom_bar(stat = "count",position = "stack")+
  theme_minimal()+
  coord_flip()


data %>%
  ->data

str(data)
data$Linked_person <- as.factor(data$Linked_person)

data%>%
  filter(!is.na(Survived))%>%
  ggplot(mapping = aes(x = Linked_person , fill = as.factor(Survived)))+
  geom_bar(stat = "count")+
  theme_minimal()+
  coord_flip()+
  facet_wrap(~Sex+age_category)



```



```{r}
library("randomForest")

data %>%
  mutate(Embarked_Pclass = paste0(Embarked,"_",Pclass),
         Embarked_Pclass = as.factor(Embarked_Pclass)) ->data

data %>%
  mutate_if(is.character,factor)->data

data$Survived <- as.factor(data$Survived)



train_p <- data[1:891,]
test_p <- data[892:1309,]
str(train_p)

names(train_p)
model_rf_evo <- randomForest(Survived ~ age_category + Embarked_Pclass + + statut + Sex  ,data = train_p)




solutions <- predict(model_rf_evo,test_p)

resutlats <- data.frame(PassengerId = test_p$PassengerId , Survived = solutions)
write.csv(resutlats ,file = "reslutats_finales+fare.csv",row.names = FALSE)
str(data)


plot(model_rf_evo)

```


# Tuning the Random Forest model 


```{r}
library("caret")


ctrl <- trainControl(method = "repeatedcv",number = 10,repeats = 10)

grid_rf <- expand.grid(.mtry = c(7,8,9))

set.seed(129)

m_rf <- train(Survived ~ age_category + Embarked + Pclass + Linked_person + statut + Sex + Fare , data = train_p ,method = "rf",trControl = ctrl,tuneGrid = grid_rf)

?train()


```


```{r}
library("C50")

sum(is.na(data$Embarked))


str(test_p)
model_tree <- C5.0(Survived ~ age_category + Embarked + Pclass + Linked_person + statut + Sex + Fare,data = train_p)

summary(model_tree)
prediction_tree <- predict(model_tree,test_p)
prediction

solutions_tree <- data.frame(PassengerId = test_p$PassengerId , Survived = prediction_tree)

write.csv(solutions_tree,file = "solutions_tree.csv",row.names = FALSE)

table(resutlats$Survived,solutions_tree$Survived)
```



Mr Joseph Bruce Ismay : OwnerShip lol


```{r}
data %>%
  filter(Fare==0)%>%
  view()

data %>%
  filter(Fare > 200)%>%
  view()

data %>%
  filter(is.na(Fare))%>%
  view()


data$Fare[is.na(data$Fare)]<- 0
```


```{r}

train_p %>%
  mutate(Embarked_Pclass = fct_reorder(Embarked_Pclass,as.numeric(Survived),mean))%>%
  ggplot(mapping = aes(x = Embarked_Pclass , fill = as.factor(Survived)) )+
  geom_bar(stat = "count")+
  theme_minimal()+
  coord_flip()+
  facet_wrap(~Sex,"free")

```

Let's try xgboost

```{r}
install.packages("xgboost")
library("xgboost")

model_mat <- model.matrix(~. , data = train_p[c("Embarked_Pclass","age_category","Sex","Fare","Linked_person","statut")])

model_test <- model.matrix(~. , data = test_p[c("Embarked_Pclass","age_category","Sex","Fare","Linked_person","statut")])

labels <- as.numeric(train_p$Survived)-1
labels_test <- test_p$Survived

dtest <- xgb.DMatrix(data=model_test,label = labels_test)
dtrain <- xgb.DMatrix(data = model_mat,label = labels) 

model <- xgboost(data = dtrain, # the data   
                 nround = 39, # max number of boosting iterations
                 objective = "binary:logistic")
solutions_xgboost<- predict(model,dtest,type= "response")
solutions_xgboost <- ifelse(solutions_xgboost>=0.5,1,0)

table(solutions_boost$Survived,resutlats$Survived)
model

solutions_boost <- data.frame(PassengerId = test_p$PassengerId , Survived = solutions_xgboost)

write.csv(solutions_boost,file = "solutions_boost.csv",row.names = FALSE)

params <- list(booster = "gbtree", objective = "binary:logistic", eta=0.3, gamma=0, max_depth=6, min_child_weight=1, subsample=1, colsample_bytree=1)

xgbcv <- xgb.cv( params = params, data = dtrain, nrounds = 100, nfold = 5, showsd = T, stratified = T, print.every.n = 10, early.stop.round = 20, maximize = F)

xgbcv$best_iteration


```

